---
output: github_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![R-CMD-check](https://github.com/castanedaM/NicheR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/castanedaM/NicheR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NicheR
### Ellipsoid-Based Virtual Species and Niche Modeling in E-space and G-space

**NicheR** is an R package for building, visualizing, predicting, and
sampling **ellipsoid-based ecological niches** using environmental data.

It provides:

- Explicit ellipsoid construction in n-dimensional environmental space
- Efficient suitability detection using Mahalanobis distance
- Flexible virtual occurrence sampling strategies (center, edge, random)
- Bias surface preparation and application for realistic sampling simulation
- Visualization tools in environmental space (E-space) and geographic space (G-space)
- A unified, fully scriptable R-native workflow

Inspired by the conceptual foundations of **NicheA** and the flexibility
of the **virtualspecies** package, **NicheR** provides a reproducible
framework for virtual species simulation and niche theory exploration.

---

## Authors

- Mariana Castaneda-Guzman
- Paanwaris Paansri
- Connor Hughes
- Marlon Cobos

---

## Installation

Install the development version from GitHub:

```{r}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("castanedaM/nicheR")
library(nicheR)
```

---

## Conceptual Framework

NicheR operates across two complementary spaces:

- **E-space (Environmental Space)**
  Ellipsoids represent multivariate environmental tolerances.

- **G-space (Geographic Space)**
  Predictions are projected across raster layers to identify suitable
  geographic regions.

This dual-space structure allows explicit separation between niche
definition, projection, and sampling processes.

---

## Workflow Overview

A typical NicheR workflow follows these steps:

1. **Prepare environmental data** — load and subset raster layers
2. **Prepare bias layers** (optional) — define sampling biases with directional effects
3. **Build an ellipsoid** — define the niche from environmental ranges
4. **Predict suitability** — project the ellipsoid onto raster or data frame inputs
5. **Apply bias** (optional) — weight predictions by a composite bias surface
6. **Sample occurrences** — draw virtual occurrence records using various strategies

---

## Quick Example

```{r}
library(nicheR)
library(terra)

# Load environmental layers
bios <- terra::rast(system.file("extdata/ma_bios.tif", package = "nicheR"))
bios <- bios[[c("bio_1", "bio_12", "bio_15")]]

bios_df <- as.data.frame(bios, xy = TRUE)
```

---

## Preparing Bias Layers

Bias layers represent factors that influence detectability or sampling effort.
Each layer can be applied in a `"direct"` or `"inverse"` direction — for example,
species richness increases sampling probability (direct), while light pollution
decreases it (inverse).

```{r}
bias <- prepare_bias(bias_surface = terra::rast(system.file("extdata/ma_biases.tif", package = "nicheR")),
                     effect_direction = c("direct", "inverse"),
                     include_processed_layers = TRUE)

# Inspect the combination formula and composite surface
bias$combination_formula
bias$composite_surface

# Visualize standardized layers and the composite bias surface
terra::plot(bias$processed_layers)
terra::plot(bias$composite_surface)
```

---

## Building an Ellipsoid

Define an ellipsoid niche from environmental ranges. At minimum, provide
a data frame with two rows (min and max) for each environmental variable.

```{r}
range <- data.frame(bio_1  = c(20, 25),
                    bio_12 = c(1000, 1500),
                    bio_15 = c(60, 75))

ell <- build_ellipsoid(range = range)

# Inspect the ellipsoid object
names(ell)
ell  # pretty print
```

### Visualizing the Ellipsoid

```{r}
# Basic plots in E-space
plot_ellipsoid(ell)
plot_ellipsoid_pairs(ell)

# With environmental background (sampled for speed)
plot_ellipsoid(ell, background = bios_df, dim = c(1, 2), bg_sample = 5000)
plot_ellipsoid(ell, background = bios_df, dim = c(1, 3), bg_sample = 5000)

# Pairwise plot with background
plot_ellipsoid_pairs(ell, background = bios_df, bg_sample = 5000, col_ell = "red")
```

---

## Predicting Suitability

Project the ellipsoid onto environmental data to obtain suitability and
Mahalanobis distance surfaces. Both truncated (inside ellipsoid only) and
non-truncated outputs are supported.

### Raster input

```{r}
ell_predict_r <- predict(ell,
                         newdata = bios,
                         suitability_truncated = TRUE,
                         mahalanobis_truncated = TRUE,
                         keep_data = TRUE   # retain original bio layers in output
                         )

terra::plot(ell_predict_r)
```

### Data frame input

```{r}
ell_predict_df <- predict(ell,
                          newdata = bios_df,
                          suitability_truncated = TRUE
                          )

head(ell_predict_df)
```

### Visualizing predictions in E-space

```{r}
# Non-truncated view
plot_ellipsoid(ell, background = ell_predict_df, pch = 20, bg_sample = 5000)
add_data(data = ell_predict_df, x = "bio_1", y = "bio_12",
         col_layer = "suitability",
         rev_pal   = TRUE,
         pal       = terrain.colors(100),
         bg_sample = 10000,
         pch       = 20)
add_ellipsoid(ell, col_ell = "red")

# Truncated view
plot_ellipsoid(ell)
add_data(data = ell_predict_df, x = "bio_1", y = "bio_12",
         col_layer = "suitability_trunc",
         rev_pal = TRUE,
         pch = 20)
add_ellipsoid(ell, col_ell = "red", lwd = 2)
```

---

## Applying Bias to Predictions

Weight suitability predictions by the composite bias surface to simulate
biased sampling conditions.

```{r}
biased_predict_r <- apply_bias(prepared_bias = bias,
                               prediction = ell_predict_r,
                               prediction_layer = "suitability")
```

---

## Sampling Virtual Occurrences

NicheR supports multiple sampling strategies and methods. Sampling can be
**unbiased** (using `sample_data()`) or **biased** (using `sample_biased_data()`).

### Unbiased sampling — raster input

```{r}
sample_data_r <- sample_data(n_occ = 100,
                             prediction = ell_predict_r,
                             prediction_layer = "suitability",
                             sampling = "centroid",
                             method = "suitability",
                             strict = TRUE)
head(sample_data_r)
```

### Unbiased sampling — data frame, centroid strategy

```{r}
sample_data_df_cn <- sample_data(n_occ = 100,
                                 prediction = ell_predict_df,
                                 prediction_layer = "suitability_trunc",
                                 sampling = "centroid",
                                 method = "mahalanobis",
                                 strict = TRUE # restrict points to inside the ellipsoid
                                 )

head(sample_data_df_cn)
```

> **Note:** Using a truncated prediction layer with `strict = TRUE` ensures
> all sampled points fall within the ellipsoid boundary.

### Unbiased sampling — edge strategy

```{r}
sample_data_df_ed <- sample_data(
  n_occ            = 100,
  prediction       = ell_predict_df,
  prediction_layer = "suitability_trunc",
  sampling         = "edge",
  method           = "suitability"
)
```

### Biased sampling — raster input

```{r}
sample_data_b <- sample_biased_data(n_occ = 100,
                                    prediction = biased_predict_r,
                                    prediction_layer = "suitability_biased_direct")

head(sample_data_b)
```

> Even with raster input, `sample_data()` and `sample_biased_data()` return a data frame of
> occurrence records, not spatial points.

---

## Visualizing Sampling Results

### Comparing centroid and edge sampling

```{r}
par(mfrow = c(2, 1))

# Centroid
plot_ellipsoid(ell, main = "Centroid Sampling — Unbiased")
add_data(data = ell_predict_df, x = "bio_1", y = "bio_12",
         pts_col = "grey", pch = 20, bg_sample = 8000)
add_data(data = sample_data_df_cn, x = "bio_1", y = "bio_12",
         pts_col = "black", pch = 4, lwd = 2)
add_ellipsoid(ell, col_ell = "red", lwd = 2)

# Edge
plot_ellipsoid(ell, main = "Edge Sampling — Unbiased", pch = 20)
add_data(data = ell_predict_df, x = "bio_1", y = "bio_12",
         pts_col = "grey", pch = 20, bg_sample = 8000)
add_data(data = sample_data_df_ed, x = "bio_1", y = "bio_12",
         pts_col = "black", pch = 4, lwd = 2)
add_ellipsoid(ell, col_ell = "red", lwd = 2)
```

---

## Contributing

We welcome contributions. If you have suggestions, bug reports, feature
requests, or code improvements, please open an issue or submit a pull
request on GitHub.

When contributing:

- Document functions using roxygen2
- Include minimal reproducible examples
- Add tests when appropriate

Thank you for helping improve **NicheR**.
