---
title: "Generate occurrence data with virtual data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate occurrence data with virtual data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 7,
  dpi = 100,
  out.width = "95%"
)
```

# Summary

-   [Description](#description)
-   [Getting ready](#sec-getting-ready)
-   [Basic generation](#sec-basic-generation)
    -   [Visualizing basic generation](#sec-visualizing-basic-generation)
    -   [1. Environmental Space Bio1 vs. Bio12](#sec-1.-environmental-space-bio1-vs.-bio12)
    -   [2. Environmental Space Bio1 vs. Bio15](#sec-2.-environmental-space-bio1-vs.-bio15)
-   [Effect of the `sampling` argument](#sec-effect-of-the-sampling-argument)
    -   [Visualizing spatial bias](#sec-visualizing-spatial-bias)
-   [Effect of the `method` argument](#sec-effect-of-the-method-argument)
    -   [Visualizing probability weights](#sec-visualizing-probability-weights)
-   [Effect of the `strict` argument](#sec-effect-of-the-strict-argument)
    -   [Visualizing boundary limits](#sec-visualizing-boundary-limits)
-   [A note on G-space](#sec-a-note-on-g-space)
-   [Save and export](#sec-save-and-export)

<hr>

# Description {#description}

Virtual species are highly useful for controlled, purely theoretical experiments in quantitative ecology. This workflow allows you to simulate species data entirely within Environmental Space (E-Space), bypassing geographic rasters. This is ideal for testing how models perform under different sampling biases without the confounding influence of real-world landscape fragmentation.

<br>

# Getting ready {#sec-getting-ready}

First, we load nicheR and define a fundamental niche. We also use virtual_data() to generate an environmental "background" of 1,000 points that our virtual species can potentially inhabit.

```{r get_ready, results='hide', message=FALSE, warning=FALSE}
# Load packages
library(nicheR)

# 1. Define Niche Space
range_df <- data.frame(bio_1  = c(22, 28),
                       bio_12 = c(1000, 3500),
                       bio_15 = c(50, 70))
ell <- build_ellipsoid(range = range_df)

# 2. Generate Environmental Background (The "Universe")
# We draw 1,000 points from the mathematical distribution of the niche
virt_bg <- virtual_data(ell, n = 1000, truncate = FALSE, effect = "direct", seed = 1)

# 3. Predict Suitability for the Background
# We score our background points so we can sample from them
pred_virt <- predict(ell, newdata = virt_bg, include_suitability = TRUE, 
                     suitability_truncated = TRUE, keep_data = TRUE)
```

<br>

# Basic generation {#sec-basic-generation}

The `sample_virtual_data()` function acts as the non-spatial twin to `sample_data()`. It picks "occurrences" from our virtual background based on the suitability scores we just calculated.

```{r}
# Generate basic virtual occurrence data
occ_virt_basic <- sample_virtual_data(
  n_occ = 100, 
  object = ell,
  virtual_prediction = pred_virt, 
  prediction_layer = "suitability", 
  seed = 123
)

head(occ_virt_basic)
```

<br>

## Visualizing basic generation {#sec-visualizing-basic-generation}

Since virtual data does not have Latitude/Longitude, we visualize it exclusively in E-Space. We will look at two pairs of dimensions to see how the generated points sit within the niche volume.

<br>

### **1. Environmental Space Bio1 vs. Bio12** {#sec-1.-environmental-space-bio1-vs.-bio12}

The gray dots are our 1,000 background points. The orange dots are the 100 individuals we sampled based on suitability. Notice they cluster toward the center (red square) because higher suitability exists there.

```{r}
plot_ellipsoid(ell, dim = c(1, 2), pch = ".", col_bg = "#9a9797", 
               xlab = "Bio1 (Temp)", ylab = "Bio12 (Precip)", main = "E-Space: Temp vs Precip")
add_data(occ_virt_basic, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
add_data(as.data.frame(t(ell$centroid)), x = "bio_1", y = "bio_12", pts_col = "red", pch = 15, cex = 1.5)
```

<br>

### **2. Environmental Space Bio1 vs. Bio15** {#sec-2.-environmental-space-bio1-vs.-bio15}

Viewing the same points along the Seasonality (Bio15) axis ensures the sampling is consistent across the 3D ellipsoid.

```{r}
plot_ellipsoid(ell, dim = c(1, 3), pch = ".", col_bg = "#9a9797", 
               xlab = "Bio1 (Temp)", ylab = "Bio15 (Seasonality)", main = "E-Space: Temp vs Seasonality")
add_data(occ_virt_basic, x = "bio_1", y = "bio_15", pts_col = "orange", pch = 20)
add_data(as.data.frame(t(ell$centroid)), x = "bio_1", y = "bio_15", pts_col = "red", pch = 15, cex = 1.5)
```

<br>

# Effect of the `sampling` argument {#sec-effect-of-the-sampling-argument}

The `sampling` argument dictates which part of the niche the individuals "prefer" to occupy.

```{r}
occ_cent <- sample_virtual_data(100, ell, pred_virt, "suitability", sampling = "centroid", seed = 123)
occ_edge <- sample_virtual_data(100, ell, pred_virt, "suitability", sampling = "edge", seed = 123)
occ_rand <- sample_virtual_data(100, ell, pred_virt, "suitability", sampling = "random", seed = 123)
```

<br>

## Visualizing spatial bias {#sec-visualizing-spatial-bias}

In virtual space, "spatial bias" refers to the distribution relative to the niche center.

**Centroid vs. Edge vs. Random:** In **Centroid**, orange dots huddle at the red square. In **Edge**, they are repelled to the purple boundary line. In **Random**, they fill the ellipsoid evenly.

```{r}
par(mfrow = c(3, 1), mar = c(4, 4, 3, 2)) 
# Centroid
plot_ellipsoid(ell, dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Centroid Sampling")
add_data(occ_cent, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
# Edge
plot_ellipsoid(ell, dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Edge Sampling")
add_data(occ_edge, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
# Random
plot_ellipsoid(ell, dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Random Sampling")
add_data(occ_rand, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

<br>

# Effect of the `method` argument {#sec-effect-of-the-method-argument}

The `method` argument changes the mathematical weight used to draw points from the 1,000 background candidates.

```{r}
occ_meth_suit <- sample_virtual_data(100, ell, pred_virt, "suitability", method = "suitability", seed = 123)
occ_meth_maha <- sample_virtual_data(100, ell, pred_virt, "Mahalanobis", method = "mahalanobis", seed = 123)
```

<br>

## Visualizing probability weights {#sec-visualizing-probability-weights}

**Suitability vs. Mahalanobis:** The **Mahalanobis** method creates a much tighter concentration of points in the niche core compared to the **Suitability** method, as it penalizes environmental distance more severely.

```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 3, 2)) 
# Suitability
plot_ellipsoid(ell, dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Method = Suitability")
add_data(occ_meth_suit, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
# Mahalanobis
plot_ellipsoid(ell, dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Method = Mahalanobis")
add_data(occ_meth_maha, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

<br>

# Effect of the `strict` argument {#sec-effect-of-the-strict-argument}

The `strict` argument determines if the virtual individuals can exist in "sink" environments outside the fundamental niche boundary.

```{r}
# Allow points in marginal background outside the ellipse
occ_lax <- sample_virtual_data(100, ell, pred_virt, "suitability", strict = FALSE, seed = 123)

# Force points to remain inside the ellipse
occ_strict <- sample_virtual_data(100, ell, pred_virt, "suitability_trunc", strict = TRUE, seed = 123)
```

<br>

## Visualizing boundary limits {#sec-visualizing-boundary-limits}

**Strict = FALSE vs. Strict = TRUE:** When `strict = FALSE`, you can see orange dots drifting outside the purple ellipse line. When `strict = TRUE`, the boundary acts as a hard wall.

```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 3, 2)) 
# False
plot_ellipsoid(ell, dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Strict = FALSE")
add_data(occ_lax, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
# True
plot_ellipsoid(ell, dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Strict = TRUE")
add_data(occ_strict, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

<br>

# A note on G-space {#sec-a-note-on-g-space}

Because the `virtual_data()` function generates points purely based on statistical distributions within environmental dimensions (e.g., Temperature and Precipitation), **these points do not possess spatial coordinates (Longitude/Latitude).**

Consequently, pure virtual species simulated via this method exist exclusively in E-Space and cannot be directly plotted onto a geographic map (G-Space) without first projecting the niche model onto a physical landscape raster.

<br>

# Save and export {#sec-save-and-export}

Since the generated virtual occurrences are standard data frames, we save them using `saveRDS()`.

```{r}
# Save the basic virtual occurrence dataset
# saveRDS(occ_virt_basic, file = "data/virtual_occurrences.rd")
```
