---
title: "Sampling occurrence data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling occurrence data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 10,
  dpi = 100,
  out.width = "95%"
)
```

# Summary

-   [Description](#description)
-   [Getting ready](#getting-ready)
-   [Basic generation](#basic-generation)
    -   [Visualizing basic generation](#visualizing-basic-generation)
        -   [1. Geographic Space](#sec-1.-geographic-space-)
        -   [2. Environmental Space Bio1 vs. Bio12](#sec-2.-environmental-space-bio1-vs.-bio12)
        -   [3. Environmental Space Bio1 vs. Bio15](#sec-3.-environmental-space-bio1-vs.-bio15)
-   [Effect of the sampling argument](#effect-of-the-sampling-argument)
    -   [Visualizing spatial bias](#visualizing-spatial-bias)
        -   [1. Centroid Sampling](#sec-1.-centroid-sampling)
        -   [2. Edge Sampling](#sec-2.-edge-sampling)
        -   [3. Random Sampling](#sec-3.-random-sampling)
-   [Effect of the method argument](#effect-of-the-method-argument)
    -   [Visualizing probability weights](#visualizing-probability-weights)
        -   [1. Suitability Method](#sec-1.-suitability-method)
        -   [2. Mahalanobis Method](#sec-2.-mahalanobis-method)
-   [Effect of the strict argument](#effect-of-the-strict-argument)
    -   [Visualizing boundary limits](#visualizing-boundary-limits)
        -   [1. Strict = FALSE](#sec-1.strict--false)
        -   [1. Strict = TRUE](#sec-1.strict--true)
-   [Save and export](#save-and-export)

<hr>

# Description {#description}

The `nicheR` package allows you to simulate how a species might be distributed across a landscape by generating synthetic occurrence data from a continuous prediction surface. Rather than viewing this as "sampling" empirical field data, the `sample_data()` function acts as a virtual ecologist, generating presence points across the available environmental background based on the species' theoretical preferences.

This vignette focuses exclusively on the `sample_data()` function. We will explore how adjusting its core arguments—`sampling`, `method`, and `strict`—alters the spatial and environmental distribution of the generated occurrences.

<br>

# Getting ready {#getting-ready}

First, we load `nicheR` and `terra`. For this vignette, we assume you have already defined a fundamental niche ellipsoid and projected it onto geographic space to create a prediction surface.

If you are unfamiliar with these prerequisite steps, please review the "**Creating Ellipsoid Based Niches"** and "**Predict"** vignettes first.

```{r get_ready, results='hide', message=FALSE, warning=FALSE}
# Load packages
library(nicheR)
library(terra)

# 1. Load environmental background
bios <- terra::rast(system.file("extdata", "ma_bios.tif", package = "nicheR"))
vars <- c("bio_1", "bio_12", "bio_15")

# 2. Load reference niche (nicheR_ellipsoid object)
data("example_sp_4", package = "nicheR")

# 3. Load pre-calculated prediction surface (from previous vignette)
# This SpatRaster contains "suitability", "Mahalanobis", "suitability_trunc", etc.
pred <- terra::rast(system.file("extdata", "predictions_3d_rast.tif", package = "nicheR"))
```

<br>

# Basic generation {#basic-generation}

The `sample_data()` function generates a spatial data frame of coordinates. By default, it requires the number of occurrences you want to generate (`n_occ`), the prediction spatial raster (`prediction`), and the specific layer to use as the probability weight (`prediction_layer`).

Let's generate a basic set of 100 occurrences using the default suitability layer.

```{r}
# Generate basic occurrence data
occ_basic <- sample_data(
  n_occ = 100,
  prediction = pred,
  prediction_layer = "suitability",
  seed = 123
)

head(occ_basic)
```

<br>

## Visualizing basic generation {#visualizing-basic-generation}

We can visualizendefined these generated points in both **Geographic Space (G-Space)** (the physical landscape) and **Environmental Space (E-Space)** (the multi-dimensional climate space). Every physical location on the map corresponds to a specific set of climate coordinates inside the fundamental niche ellipsoid.

<br>

### **1. Geographic Space** {#sec-1.-geographic-space-}

First, let's look at the physical map. The background colors represent habitat suitability (green is high, brown/white is low). The black dots are the physical coordinates (Longitude/Latitude) where our 100 virtual individuals were generated. Because we generated points based on suitability, notice how the dots cluster in the greenest areas.

```{r}
terra::plot(pred[["suitability"]], main = "G-Space: Geographic Map")
points(occ_basic[, c("x", "y")], pch = 20, col = "red", cex = 1.2)
```

<br>

### **2. Environmental Space Bio1 vs. Bio12** {#sec-2.-environmental-space-bio1-vs.-bio12}

Now, let's shift our view from geography to climate. Here, we plot the exact same 100 individuals, but instead of mapping their physical location, we map the climate they are experiencing. The gray dust represents the available background climate in our study area. The large ellipse is the fundamental niche boundary. The **red square** is the centroid—the "perfect" climate for the species. The **orange dots** are our generated occurrences.

```{r}
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 2), pch = ".", col_bg = "#9a9797", 
               xlab = "Bio1 (Temp)", ylab = "Bio12 (Precip)", main = "E-Space: Temp vs Precip")
add_data(occ_basic, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
add_data(as.data.frame(t(example_sp_4$centroid)), x = "bio_1", y = "bio_12", pts_col = "red", pch = 15, cex = 1.5)
```

<br>

### **3. Environmental Space Bio1 vs. Bio15** {#sec-3.-environmental-space-bio1-vs.-bio15}

Because niches are multi-dimensional, we must look at other axes. Here, we keep Temperature (Bio1) on the X-axis but change the Y-axis to Precipitation Seasonality (Bio15) to see how the species tolerates fluctuations in rainfall.

```{r}
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 3), pch = ".", col_bg = "#9a9797", 
               xlab = "Bio1 (Temp)", ylab = "Bio15 (Seasonality)", main = "E-Space: Temp vs Seasonality")
add_data(occ_basic, x = "bio_1", y = "bio_15", pts_col = "orange", pch = 20)
add_data(as.data.frame(t(example_sp_4$centroid)), x = "bio_1", y = "bio_15", pts_col = "red", pch = 15, cex = 1.5)
```

By default, the function generates points proportional to habitat suitability (the niche centroid). Therefore, you can see the orange dots clumping somewhat near the red centroid in E-Space, which translates to black dots clumping in the greenest areas of the G-Space map.

<br>

# Effect of the `sampling` argument {#effect-of-the-sampling-argument}

The `sampling` argument determines the spatial bias of the selection probability. It answers the question: *Where within the niche is the species most likely to be found?*

-   **`centroid`**: The species strongly prefers its optimal conditions. Generates points clustered near the center of the niche.

-   **`edge`**: The species is frequently pushed into marginal environments (e.g., due to competition). Generates points scattered toward the niche boundaries.

-   **`random`**: The species is distributed uniformly across all suitable habitats, regardless of optimality.

```{r}
occ_cent <- sample_data(100, pred, "suitability", sampling = "centroid", seed = 123)
occ_edge <- sample_data(100, pred, "suitability", sampling = "edge", seed = 123)
occ_rand <- sample_data(100, pred, "suitability", sampling = "random", seed = 123)
```

<br>

## Visualizing spatial bias {#visualizing-spatial-bias}

Let's compare these three strategies side-by-side in both G-Space and E-Space.

<br>

### **1. Centroid Sampling** {#sec-1.-centroid-sampling}

Notice how the orange dots huddle tightly around the red square in E-Space. Geographically, this translates to highly concentrated, clumped points on the map in only the most pristine habitats.

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 2)) 
terra::plot(pred[["suitability"]], main = "G-Space: Centroid Sampling"); points(occ_cent[, 1:2], pch = 20, col = "red")
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Centroid Sampling")
add_data(occ_cent, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

<br>

### **2. Edge Sampling** {#sec-2.-edge-sampling}

Here, the orange dots are repelled from the center and hug the outer boundary of the ellipse. Geographically, this forces the species into fragmented, peripheral habitats on the fringes of its range.

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 2)) 
terra::plot(pred[["suitability"]], main = "G-Space: Edge Sampling"); points(occ_edge[, 1:2], pch = 20, col = "red")
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Edge Sampling")
add_data(occ_edge, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

<br>

### **3. Random Sampling** {#sec-3.-random-sampling}

The points are scattered broadly across all suitable environments without bias toward the core or the edge.

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 2)) 
terra::plot(pred[["suitability"]], main = "G-Space: Random Sampling"); points(occ_rand[, 1:2], pch = 20, col = "red")
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Random Sampling")
add_data(occ_rand, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

<br>

# Effect of the `method` argument {#effect-of-the-method-argument}

The `method` argument controls the underlying mathematical weight used to draw the samples.

-   **`suitability`**: Weights the probability of occurrence linearly based on the 0-1 habitat suitability index.

-   **`mahalanobis`**: Weights the probability based on the multivariate environmental distance from the centroid. This often creates a more dramatic, non-linear drop-off in generation probability compared to suitability.

```{r}
occ_meth_suit <- sample_data(100, pred, "suitability", method = "suitability", seed = 123)
occ_meth_maha <- sample_data(100, pred, "Mahalanobis", method = "mahalanobis", seed = 123)
```

<br>

## Visualizing probability weights {#visualizing-probability-weights}

<br>

### **1. Suitability Method** {#sec-1.-suitability-method}

Points are drawn linearly based on habitat quality.

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 2)) 
terra::plot(pred[["suitability"]], main = "G-Space: Suitability Method"); points(occ_meth_suit[, 1:2], pch = 20, col = "red")
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Suitability Method")
add_data(occ_meth_suit, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

<br>

### **2. Mahalanobis Method** {#sec-2.-mahalanobis-method}

Because Mahalanobis distance penalizes deviations from the centroid heavily, generating points based on this method creates an even stricter concentration of points in optimal environments. This is highly useful for simulating sensitive, specialist species.

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 2)) 
terra::plot(pred[["Mahalanobis"]], main = "G-Space: Mahalanobis Method"); points(occ_meth_maha[, 1:2], pch = 20, col = "red")
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Mahalanobis Method")
add_data(occ_meth_maha, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

<br>

# Effect of the `strict` argument {#effect-of-the-strict-argument}

By default, generating occurrences allows points to fall slightly outside the strict boundaries of the fundamental niche ellipsoid (`strict = FALSE`). This is ecologically realistic, as it simulates sink populations, source-sink dynamics, or the influence of unmeasured micro-climates.

Setting `strict = TRUE` (and using a truncated prediction layer) establishes a hard boundary, forbidding any generated points from falling in unsuitable environments.

```{r}
# strict = FALSE (Allows points outside the ellipse)
occ_lax <- sample_data(100, pred, "suitability", strict = FALSE, seed = 123)

# strict = TRUE (Restricts points strictly to the ellipse using truncated layers)
occ_strict <- sample_data(100, pred, "suitability_trunc", strict = TRUE, seed = 123)
```

<br>

## Visualizing boundary limits {#visualizing-boundary-limits}

### **1. Strict = FALSE** {#sec-1.strict--false}

In the E-Space plot below, notice how a few rogue orange dots fall just outside the purple boundary line of the ellipsoid. These represent our simulated sink populations.

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 2)) 
terra::plot(pred[["suitability"]], main = "G-Space: Strict = FALSE"); points(occ_lax[, 1:2], pch = 20, col = "red")
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Strict = FALSE")
add_data(occ_lax, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
```

### **2. Strict = TRUE**

Here, the generation algorithm respects the hard mathematical threshold—every single orange dot is contained strictly within the boundary line.

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 2)) 
terra::plot(pred[["suitability_trunc"]], main = "G-Space: Strict = TRUE"); points(occ_strict[, 1:2], pch = 20, col = "red")
plot_ellipsoid(example_sp_4, background = as.data.frame(bios[[vars]]), dim = c(1, 2), pch = ".", col_bg = "#9a9797", main = "E-Space: Strict = TRUE")
add_data(occ_strict, x = "bio_1", y = "bio_12", pts_col = "orange", pch = 20)
dev.off()
```

<br>

# Save and export {#save-and-export}

Once you are satisfied with your generated occurrence points, you can easily save the resulting data frame for use in other software or later R sessions. We will save the `occ_basic` dataset we generated at the very beginning of this vignette.

```{r}
# Save the basic generated occurrence data frame to your local directory
saveRDS(occ_basic, file = "data/virtual_occurrences.rds")

# To load this data back into a future session:
# my_occurrences <- readRDS("data/virtual_occurrences.rds")
```
