---
title: "Visualizing ellipsoids in environmental space"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing ellipsoids in environmental space}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  dpi = 100,
  out.width = "95%"
)
```

## Summary

- [Description](#description)
- [Getting ready](#getting-ready)
- [Loading example data](#loading-example-data)
- [plot_ellipsoid()](#plot_ellipsoid)
  - [Ellipsoid boundary only](#ellipsoid-boundary-only)
  - [Background points](#background-points)
  - [Prediction colored by a continuous variable](#prediction-colored-by-a-continuous-variable)
  - [Truncated predictions and grey outside region](#truncated-predictions-and-grey-outside-region)
  - [Reversed palette and transparency](#reversed-palette-and-transparency)
  - [Subsampling large datasets](#subsampling-large-datasets)
  - [Fixed axis limits for comparing ellipsoids](#fixed-axis-limits-for-comparing-ellipsoids)
- [add_data() and add_ellipsoid()](#add_data-and-add_ellipsoid)
  - [Layering background, records, and boundary](#layering-background-records-and-boundary)
  - [Coloring records by a continuous variable](#coloring-records-by-a-continuous-variable)
- [plot_ellipsoid_pairs()](#plot_ellipsoid_pairs)
  - [Pairs with background](#pairs-with-background)
  - [Pairs with predictions](#pairs-with-predictions)

<hr>

## Description

This vignette demonstrates the plotting functions available in `nicheR` for visualizing ellipsoidal niche models in environmental space (E-space). All plots operate in E-space: axes represent environmental variables and points represent environmental combinations from background data or predictions.

The functions covered are:

- `plot_ellipsoid()`: opens a new E-space plot showing the ellipsoid boundary, optionally with background points or a prediction surface.
- `add_data()`: overlays points onto an existing plot. Points can be a single color or colored by a continuous variable such as suitability.
- `add_ellipsoid()`: overlays an ellipsoid boundary onto an existing plot.
- `plot_ellipsoid_pairs()`: produces a multi-panel layout showing all pairwise two-dimensional projections of an ellipsoid, with globally consistent axis limits.

These functions are designed to be layered. The typical workflow is to open a plot with `plot_ellipsoid()` and then add data or boundaries on top with `add_data()` and `add_ellipsoid()`.

<br>

## Getting ready

If `nicheR` has not been installed yet, please do so. See the [Main guide](index.html) for installation instructions.

```{r get_ready, results='hide', message=FALSE, warning=FALSE}
library(nicheR)

# Saving original plotting parameters
original_par <- par(no.readonly = TRUE)
```

<br>

## Loading example data

We use the same example data as in the other vignettes: a two-dimensional `nicheR_ellipsoid` object for a reference species and the corresponding North American background data.

```{r data}
data("ref_ellipse", package = "nicheR")
data("back_data",   package = "nicheR")
```

<br>

We will also need some predictions to demonstrate the prediction-based plotting modes. We compute them all here so they are available throughout the vignette.

```{r predictions, message=FALSE}
# Non-truncated Mahalanobis distance
pred_maha <- predict(ref_ellipse,
                     newdata = back_data[, ref_ellipse$var_names],
                     include_mahalanobis = TRUE,
                     include_suitability = FALSE,
                     verbose = FALSE)

# Non-truncated suitability
pred_suit <- predict(ref_ellipse,
                     newdata = back_data[, ref_ellipse$var_names],
                     include_mahalanobis = FALSE,
                     include_suitability = TRUE,
                     verbose = FALSE)

# Truncated suitability (outside = 0, inside = suitability value)
pred_trunc <- predict(ref_ellipse,
                      newdata = back_data[, ref_ellipse$var_names],
                      include_mahalanobis = FALSE,
                      include_suitability = FALSE,
                      suitability_truncated = TRUE,
                      verbose = FALSE)
```

```{r include=FALSE}
# Shared settings used throughout
mars     <- c(4, 4, 2, 1)
blue_pal <- hcl.colors(100, palette = "Oslo", rev = TRUE)
vir_pal  <- hcl.colors(100, palette = "Viridis")
```

<br>

## plot_ellipsoid()

`plot_ellipsoid()` opens a new E-space plot. It has three display modes depending on what is provided:

1. **Ellipsoid only**: neither `background` nor `prediction` is given. The function draws just the ellipsoid boundary, sized to the boundary itself. This is the starting point for building layered plots manually with `add_data()` and `add_ellipsoid()`.
2. **Background**: a `background` data frame is given. Background points are drawn first, the ellipsoid boundary is drawn on top, and axis limits span both.
3. **Prediction**: a `prediction` data frame is given. When `col_layer` is also specified, points are colored by that column and the axis limits span the full prediction extent.

<br>

### Ellipsoid boundary only

When neither `background` nor `prediction` is given, `plot_ellipsoid()` draws only the ellipsoid boundary. The plot is sized to the boundary extent, so axis limits reflect only the ellipsoid dimensions. This mode is useful when you want full manual control over what goes into the plot.

```{r boundary_only}
par(mar = mars)

plot_ellipsoid(ref_ellipse,
               col_ell = "#e10000",
               lwd = 2,
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Ellipsoid boundary only")
```

<br>

### Background points

Passing a `background` data frame draws the background points first and the ellipsoid boundary on top. Axis limits are computed from both the background extent and the ellipsoid boundary, so the ellipsoid is never clipped.

```{r background}
par(mar = mars)

plot_ellipsoid(ref_ellipse,
               background = back_data[, ref_ellipse$var_names],
               col_ell = "#e10000",
               col_bg  = "#9a9797",
               lwd = 2, pch = ".",
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Background points")
```

<br>

### Prediction colored by a continuous variable

Passing a `prediction` data frame and specifying `col_layer` colors the points by the values in that column. Any column in the prediction data frame can be used, including Mahalanobis distance, suitability, or truncated versions of either.

For non-truncated predictions every point receives a color from the palette, because every point has a finite positive value. The ellipsoid boundary is drawn on top in `col_ell`.

```{r pred_maha}
par(mar = mars)

plot_ellipsoid(ref_ellipse,
               prediction = pred_maha,
               col_layer  = "Mahalanobis",
               pal        = blue_pal,
               col_ell    = "#e10000",
               lwd = 2, pch = 16, cex_bg = 0.5,
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Mahalanobis distance (non-truncated)")

legend("topright",
       legend = c("Ellipsoid boundary", "Low D\u00b2", "High D\u00b2"),
       col    = c("#e10000", blue_pal[5], blue_pal[90]),
       pch    = c(NA, 16, 16), lty = c(1, NA, NA),
       lwd    = c(2, NA, NA), cex = 0.8, bty = "n")
```

<br>

```{r pred_suit}
par(mar = mars)

plot_ellipsoid(ref_ellipse,
               prediction = pred_suit,
               col_layer  = "suitability",
               pal        = vir_pal,
               col_ell    = "#e10000",
               lwd = 2, pch = 16, cex_bg = 0.5,
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Suitability (non-truncated)")

legend("topright",
       legend = c("Ellipsoid boundary", "Low suitability", "High suitability"),
       col    = c("#e10000", vir_pal[5], vir_pal[95]),
       pch    = c(NA, 16, 16), lty = c(1, NA, NA),
       lwd    = c(2, NA, NA), cex = 0.8, bty = "n")
```

<br>

### Truncated predictions and grey outside region

When `col_layer` points to a truncated prediction column (`suitability_trunc` or `Mahalanobis_trunc`), `plot_ellipsoid()` automatically detects outside points, those with value zero or `NA`, and draws them in `col_bg`. Inside points are drawn on top in their palette color. Axis limits are computed from the full prediction extent so the view does not collapse to the ellipsoid interior.

```{r pred_trunc}
par(mar = mars)

plot_ellipsoid(ref_ellipse,
               prediction = pred_trunc,
               col_layer  = "suitability_trunc",
               pal        = vir_pal,
               col_bg     = "#d4d4d4",
               col_ell    = "#e10000",
               lwd = 2, pch = 16, cex_bg = 0.5,
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Truncated suitability")

legend("topright",
       legend = c("Ellipsoid boundary", "Low suitability",
                  "High suitability", "Outside (zero)"),
       col    = c("#e10000", vir_pal[5], vir_pal[95], "#d4d4d4"),
       pch    = c(NA, 16, 16, 16), lty = c(1, NA, NA, NA),
       lwd    = c(2, NA, NA, NA), cex = 0.8, bty = "n")
```

<br>

The same pattern applies to truncated Mahalanobis distance, where outside points receive `NA` and are drawn in `col_bg`.

```{r pred_trunc_maha}
pred_trunc_maha <- predict(ref_ellipse,
                           newdata = back_data[, ref_ellipse$var_names],
                           include_mahalanobis   = FALSE,
                           include_suitability   = FALSE,
                           mahalanobis_truncated = TRUE,
                           verbose = FALSE)

par(mar = mars)

plot_ellipsoid(ref_ellipse,
               prediction = pred_trunc_maha,
               col_layer  = "Mahalanobis_trunc",
               pal        = blue_pal,
               col_bg     = "#d4d4d4",
               col_ell    = "#e10000",
               lwd = 2, pch = 16, cex_bg = 0.5,
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Truncated Mahalanobis distance")

legend("topright",
       legend = c("Ellipsoid boundary", "Inside (low D\u00b2)",
                  "Inside (high D\u00b2)", "Outside (NA)"),
       col    = c("#e10000", blue_pal[5], blue_pal[90], "#d4d4d4"),
       pch    = c(NA, 16, 16, 16), lty = c(1, NA, NA, NA),
       lwd    = c(2, NA, NA, NA), cex = 0.8, bty = "n")
```

<br>

### Reversed palette and transparency

`rev_pal = TRUE` reverses the color palette so high values map to the dark end instead of the bright end. `alpha_bg` controls transparency of the data points and `alpha_ell` controls transparency of the ellipsoid boundary line, both on a 0 to 1 scale.

```{r rev_pal}
par(mar = mars)

plot_ellipsoid(ref_ellipse,
               prediction = pred_suit,
               col_layer  = "suitability",
               pal        = vir_pal,
               rev_pal    = TRUE,
               alpha_bg   = 0.5,
               alpha_ell  = 0.8,
               col_ell    = "#0004d5",
               lwd = 2, pch = 16, cex_bg = 0.5,
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Reversed palette + transparency")

legend("topright",
       legend = c("Ellipsoid boundary",
                  "Low suitability (now bright)",
                  "High suitability (now dark)"),
       col    = c("#0004d5", vir_pal[95], vir_pal[5]),
       pch    = c(NA, 16, 16), lty = c(1, NA, NA),
       lwd    = c(2, NA, NA), cex = 0.8, bty = "n")
```

<br>

### Subsampling large datasets

The `bg_sample` argument draws a random subsample before plotting. This is useful when plotting hundreds of thousands of background points would slow rendering without adding visual information. The subsample is drawn independently for inside and outside points when `col_layer` is active, so both groups remain represented.

```{r bg_sample, fig.width=7, fig.height=3.5, out.width="95%"}
par(mfrow = c(1, 2), cex = 0.75, mar = mars)

plot_ellipsoid(ref_ellipse,
               background = back_data[, ref_ellipse$var_names],
               col_ell = "#e10000", col_bg = "#9a9797",
               lwd = 2, pch = ".",
               xlab = "Bio1", ylab = "Bio12",
               main = "Full background")

plot_ellipsoid(ref_ellipse,
               background = back_data[, ref_ellipse$var_names],
               bg_sample  = 500,
               col_ell = "#e10000", col_bg = "#9a9797",
               lwd = 2, pch = 16, cex_bg = 0.4,
               xlab = "Bio1", ylab = "Bio12",
               main = "Subsampled (n = 500)")
```

<br>

### Fixed axis limits for comparing ellipsoids

The `fixed_lims` argument accepts a named list with `xlim` and `ylim` vectors and overrides the automatically computed axis limits. This is most useful when placing two ellipsoids side by side and wanting both panels on the same scale so their sizes are visually comparable.

```{r fixed_lims, fig.width=7, fig.height=3.5, out.width="95%"}
# Create a second ellipsoid with a shifted centroid for comparison
ref_ellipse2          <- ref_ellipse
ref_ellipse2$centroid <- ref_ellipse$centroid + c(5, 100)

# Define shared limits from the full background extent
global_lims <- list(
  xlim = range(back_data$bio_1,  na.rm = TRUE),
  ylim = range(back_data$bio_12, na.rm = TRUE)
)

par(mfrow = c(1, 2), cex = 0.75, mar = mars)

plot_ellipsoid(ref_ellipse,
               background = back_data[, ref_ellipse$var_names],
               fixed_lims = global_lims,
               col_ell = "#e10000", col_bg = "#9a9797",
               lwd = 2, pch = ".",
               xlab = "Bio1", ylab = "Bio12",
               main = "Ellipsoid 1")

plot_ellipsoid(ref_ellipse2,
               background = back_data[, ref_ellipse2$var_names],
               fixed_lims = global_lims,
               col_ell = "#0004d5", col_bg = "#9a9797",
               lwd = 2, pch = ".",
               xlab = "Bio1", ylab = "Bio12",
               main = "Ellipsoid 2 (shifted centroid)")
```

<br>

Without `fixed_lims`, each panel would rescale to its own data extent, making the two ellipsoids appear the same size regardless of their actual difference in position or spread.

<br>

## add_data() and add_ellipsoid()

`add_data()` and `add_ellipsoid()` layer content onto an existing plot opened by `plot_ellipsoid()`. The key thing to keep in mind is drawing order: whatever is drawn last sits on top. The typical order is background first, records second, ellipsoid boundary last.

<br>

### Layering background, records, and boundary

This is the most common layering pattern. We open the plot with a muted background, add records in a distinctive color, then draw the ellipsoid boundary on top so it is not obscured by the points.

```{r layering}
# Simulate records from inside the ellipsoid
set.seed(42)
rec_idx <- sample(which(pred_trunc$suitability_trunc > 0), 40)
rec_pts <- back_data[rec_idx, ref_ellipse$var_names]

par(mar = mars)

# Step 1: background (muted color, boundary also muted so it sits behind)
plot_ellipsoid(ref_ellipse,
               background = back_data[, ref_ellipse$var_names],
               col_bg  = "#c5c5c5",
               col_ell = "#c5c5c5",
               pch = ".", lwd = 1,
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Background + records + boundary")

# Step 2: records on top of background
add_data(rec_pts,
         x = "bio_1", y = "bio_12",
         pts_col = "#0004d5", pch = 19, cex = 1.2)

# Step 3: ellipsoid boundary on top of everything
add_ellipsoid(ref_ellipse, col_ell = "#e10000", lwd = 2)

legend("topright",
       legend = c("Background", "Records", "Ellipsoid boundary"),
       col    = c("#c5c5c5", "#0004d5", "#e10000"),
       pch    = c(16, 19, NA), lty = c(NA, NA, 1),
       lwd    = c(NA, NA, 2), cex = 0.8, bty = "n")
```

<br>

### Coloring records by a continuous variable

`add_data()` also supports `col_layer`, which maps a continuous column onto a palette the same way `plot_ellipsoid()` does. This is useful for showing suitability, Mahalanobis distance, or any other variable at the record locations on top of a background.

```{r add_data_col_layer}
# Predict suitability at the records
rec_pred <- predict(ref_ellipse,
                    newdata = rec_pts,
                    include_suitability = TRUE,
                    include_mahalanobis = FALSE,
                    verbose = FALSE)

par(mar = mars)

# Background in grey
plot_ellipsoid(ref_ellipse,
               background = back_data[, ref_ellipse$var_names],
               col_bg  = "#d4d4d4",
               col_ell = "#e10000",
               pch = ".", lwd = 2,
               xlab = "Bio1 (Mean Annual Temperature)",
               ylab = "Bio12 (Annual Precipitation)",
               main = "Records colored by suitability")

# Records colored by suitability
add_data(rec_pred,
         x = "bio_1", y = "bio_12",
         col_layer = "suitability",
         pal       = vir_pal,
         pch = 19, cex = 1.4)

add_ellipsoid(ref_ellipse, col_ell = "#e10000", lwd = 2)

legend("topright",
       legend = c("Background", "Ellipsoid boundary",
                  "Low suitability", "High suitability"),
       col    = c("#d4d4d4", "#e10000", vir_pal[5], vir_pal[95]),
       pch    = c(16, NA, 19, 19), lty = c(NA, 1, NA, NA),
       lwd    = c(NA, 2, NA, NA), cex = 0.8, bty = "n")
```

<br>

## plot_ellipsoid_pairs()

`plot_ellipsoid_pairs()` automatically generates all pairwise two-dimensional projections of an ellipsoid and arranges them in a grid. The number of panels is $\binom{p}{2}$ where $p$ is the number of dimensions. For a two-dimensional ellipsoid there is one pair; for three dimensions there are three; and so on.

The key feature is that axis limits are pre-computed globally across all variables and shared across all panels. Without this, each panel would rescale to its own data extent, making niche widths appear identical across dimensions even when they differ substantially.

<br>

### Pairs with background

```{r pairs_background, fig.width=7, fig.height=4, out.width="95%"}
# Build a 3D ellipsoid for a more interesting pairs example
range_3d <- data.frame(bio_1  = c(27, 35),
                       bio_12 = c(1000, 1500),
                       bio_15 = c(60, 75))

ellipse_3d <- build_ellipsoid(ranges = range_3d)

par(cex = 0.7)

plot_ellipsoid_pairs(ellipse_3d,
                     background = back_data[, ellipse_3d$var_names],
                     col_ell = "#e10000",
                     col_bg  = "#9a9797",
                     lwd = 2, pch = ".")
```

<br>

Each panel shows the ellipsoid projected onto a different pair of variables. Because axis limits are shared globally, the relative spread of the ellipsoid across dimensions is visually preserved. The empty cell in the 2x2 grid is a consequence of having three pairs and a layout that rounds up to four cells.

<br>

### Pairs with predictions

`plot_ellipsoid_pairs()` also accepts `prediction` directly. This is the recommended way to show suitability or Mahalanobis distance across all pairwise projections at once.

```{r pairs_prediction, fig.width=7, fig.height=4, out.width="95%"}
suit_3d <- predict(ellipse_3d,
                   newdata = back_data[, ellipse_3d$var_names],
                   include_mahalanobis   = FALSE,
                   include_suitability   = TRUE,
                   suitability_truncated = TRUE,
                   verbose = FALSE)

par(cex = 0.7)

plot_ellipsoid_pairs(ellipse_3d,
                     prediction = suit_3d,
                     col_layer  = "suitability_trunc",
                     pal        = vir_pal,
                     col_bg     = "#d4d4d4",
                     col_ell    = "#e10000",
                     lwd = 2, pch = 16, cex_bg = 0.3)
```

<br>

The grey outside region and colored inside region behave identically to a single `plot_ellipsoid()` call with a truncated `col_layer`, applied consistently across every panel.

<br>

```{r par_reset}
par(original_par)
```
