---
title: "Sampling Occurrence Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling Occurrence Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  dpi = 100,
  out.width = "95%"
)
```

## Summary

- [Description](#description)
- [Getting ready](#getting-ready)
- [1. Defining the Niche Space: build_ellipsoid()](#1-defining-the-niche-space-build_ellipsoid)
  - [Example: Building an Ellipsoid](#example-building-an-ellipsoid)
- [2. Generating the Prediction Surface](#2-generating-the-prediction-surface)
  - [Example: Predicting to a Spatial Raster](#example-predicting-to-a-spatial-raster)
- [3. Standard Occurrence Sampling: sample_data()](#3-standard-occurrence-sampling-sample_data)
  - [Example: Sampling from a Suitability Raster](#example-sampling-from-a-suitability-raster)
  - [Suitability | centroid | Strict = FALSE](#suitability-centroid-strict-false)
  - [Mahalanobis | edge | Strict = FALSE](#mahalanobis-edge-strict-false)
  - [Suitability Truncated | random | Strict = TRUE](#suitability-truncated-random-strict-true)
  - [Mahalanobis Truncated | edge | Strict = TRUE](#mahalanobis-truncated-edge-strict-true)
- [4. Introducing Sampling Bias: sample_biased_data()](#4-introducing-sampling-bias-sample_biased_data)
  - [Prerequisites: Prepare the Bias Layer](#prerequisites-prepare-the-bias-layer)
  - [Suitability | Direct Bias | Strict = FALSE](#suitability-direct-bias-strict-false)
  - [Suitability Truncated | Inverse Bias | Strict = TRUE](#suitability-truncated-inverse-bias-strict-true)
  - [Mahalanobis | Direct Bias | Strict = FALSE](#mahalanobis-direct-bias-strict-false)
  - [Mahalanobis Truncated | Inverse Bias | Strict = TRUE](#mahalanobis-truncated-inverse-bias-strict-true)
- [5. Virtual Species Sampling: sample_virtual_data()](#5-virtual-species-sampling-sample_virtual_data)
  - [Example: Sampling from a Virtual Species Object](#example-sampling-from-a-virtual-species-object)
  - [Suitability | Centroid | Strict = FALSE](#suitability-centroid-strict-false-1)
  - [Suitability Truncated | Random | Strict = TRUE](#suitability-truncated-random-strict-true-1)
  - [Mahalanobis | Edge | Strict = FALSE](#mahalanobis-edge-strict-false-1)
  - [Mahalanobis Truncated | Centroid | Strict = TRUE](#mahalanobis-truncated-centroid-strict-true)

<hr>

# Description

The nicheR package provides a flexible suite of sampling functions for drawing occurrence points from raster layers, data frames, or virtual species objects.

This vignette covers the complete core workflow:

1.  Defining the niche space (`build_ellipsoid()`).

2.  Predicting habitat suitability spatially (`predict()`).

3.  Sampling standard occurrences (`sample_data()`).

4.  Simulating biased sampling efforts (`sample_biased_data()`).

5.  Directly sampling from virtual species objects (`sample_virtual_data()`).

<br>

# Getting ready

```{r}
# Load packages
library(nicheR)
library(terra)
```

<br>

# 1. Defining the Niche Space: `build_ellipsoid()`

## Example: Building an Ellipsoid

```{r}
bios_file <- system.file("extdata", "ma_bios.tif", package = "nicheR")
bios <- rast(bios_file)
vars <- c("bio_1", "bio_12", "bio_15")

# Define environmental ranges
range_df <- data.frame(bio_1  = c(22, 28),
                       bio_12 = c(1000, 3500),
                       bio_15 = c(50, 70))

# Build ellipsoid niche
ell <- build_ellipsoid(range = range_df)
```

<br>

# 2. Generating the Prediction Surface

## Example: Predicting to a Spatial Raster

```{r}
# Predict suitability
pred <- predict(ell,
                newdata = bios,
                include_mahalanobis = TRUE,
                include_suitability = TRUE,
                suitability_truncated = TRUE,
                mahalanobis_truncated = TRUE)
plot(pred)
```

<br>

# 3. Standard Occurrence Sampling: sample_data()

**Objective:** To generate presence points from a continuous prediction surface (e.g., habitat suitability or Mahalanobis distance) using weighted random sampling.

**Purpose:** This method is essential for creating true presence datasets from a known truth to evaluate ENM predictive performance.

The sample_data() function allows you to sample points based on suitability or distance metrics. You can control the sampling direction using the sampling argument:

"centroid": Samples more frequently from areas of high suitability or low distance.

"edge": Samples more frequently from areas of low suitability or high distance.

"random": Samples uniformly across the available space.

## Example: Sampling from a Suitability Raster

## Suitability | centroid | Strict = FALSE

```{r}
# --- Sampling ---
occ_suit_cent <- sample_data(
  n_occ = 100,
  prediction = pred,
  prediction_layer = "suitability",
  sampling = "centroid",
  method = "suitability",
  seed = 123,
  strict = FALSE
)

occ_suit_cent_env <- terra::extract(bios, 
                                    occ_suit_cent[, c(1:2)]) |> 
  cbind(occ_suit_cent[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 

# Plot 1: Bio1 vs Bio12
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(1, 2),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1", 
               ylab = "Bio12",
               main = "Suitability | Centroid (Bio1 v Bio12)")
add_data(occ_suit_cent_env, 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "red",
         pch = 15)

# Plot 2: Bio12 vs Bio15
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(2, 3),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15",
               main = "Suitability | Centroid (Bio12 v Bio15)")
add_data(occ_suit_cent_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12", 
         y = "bio_15",
         pts_col = "red", 
         pch = 15)

# Mahalanobis | edge | Strict = FALSE
```

<br>

## Mahalanobis | edge | Strict = FALSE

```{r}
# --- Sampling ---
occ_maha_edge <- sample_data(
  n_occ = 100,
  prediction = pred,
  prediction_layer = "Mahalanobis",
  sampling = "edge",
  method = "mahalanobis",
  seed = 123,
  strict = FALSE
)

occ_maha_edge_env <- terra::extract(bios, occ_maha_edge[, c(1:2)]) |> 
  cbind(occ_maha_edge[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 

# Plot 1: Bio1 vs Bio12
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(1, 2),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1", 
               ylab = "Bio12",
               main = "Mahalanobis | Edge (Bio1 v Bio12)")
add_data(occ_maha_edge_env, 
         x = "bio_1",
         y = "bio_12", 
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "red",
         pch = 15)

# Plot 2: Bio12 vs Bio15
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(2, 3),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15",
               main = "Mahalanobis | Edge (Bio12 v Bio15)")
add_data(occ_maha_edge_env, 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "red", 
         pch = 15)


```

<br>

## Suitability Truncated | random | Strict = TRUE

```{r}
# --- Sampling ---
occ_suit_trunc_rand <- sample_data(
  n_occ = 100,
  prediction = pred,
  prediction_layer = "suitability_trunc",
  sampling = "random",
  method = "suitability",
  seed = 123,
  strict = TRUE
)

occ_suit_trunc_rand_env <- terra::extract(bios, 
                                          occ_suit_trunc_rand[, c(1:2)]) |> 
  cbind(occ_suit_trunc_rand[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 

# Plot 1: Bio1 vs Bio12
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(1, 2),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1", 
               ylab = "Bio12",
               main = "Suit Truncated | Random (Bio1 v Bio12)")
add_data(occ_suit_trunc_rand_env, 
         x = "bio_1", 
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "red",
         pch = 15)

# Plot 2: Bio12 vs Bio15
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(2, 3),
               pch = ".", 
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Suit Truncated | Random (Bio12 v Bio15)")
add_data(occ_suit_trunc_rand_env, 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "red", 
         pch = 15)


```

<br>

## Mahalanobis Truncated | edge | Strict = TRUE

```{r}
# --- Sampling ---
occ_maha_trunc_cent <- sample_data(
  n_occ = 100,
  prediction = pred,
  prediction_layer = "Mahalanobis_trunc",
  sampling = "edge",
  method = "mahalanobis",
  seed = 123,
  strict = TRUE
)

occ_maha_trunc_cent_env <- terra::extract(bios, 
                                          occ_maha_trunc_cent[, c(1:2)]) |> 
  cbind(occ_maha_trunc_cent[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 

# Plot 1: Bio1 vs Bio12
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(1, 2),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Maha Truncated | Centroid (Bio1 v Bio12)")
add_data(occ_maha_trunc_cent_env, 
         x = "bio_1",
         y = "bio_12", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "red", 
         pch = 15)

# Plot 2: Bio12 vs Bio15
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15",
               main = "Maha Truncated | Centroid (Bio12 v Bio15)")
add_data(occ_maha_trunc_cent_env, 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "red", 
         pch = 15)

dev.off()
```

<br>

# 4. Introducing Sampling Bias: sample_biased_data()

**Objective:** To simulate spatially biased occurrence data from a prediction layer.

**Purpose:** Real-world biodiversity data is rarely collected systematically. This function allows researchers to simulate spatial sampling bias (e.g., points collected near roads or research stations) to test how robust their ecological models are to imperfect detection and biased collection efforts.

*Note: This function applies a spatial restriction or weighting scheme to simulate uneven sampling effort.*

## Prerequisites: Prepare the Bias Layer

```{r}
# Load the bias layer
biases_file <- system.file("extdata", "ma_biases.tif", package = "nicheR")
biases <- rast(biases_file)

# Prepare bias
prep_biases <- prepare_bias(
  bias_surface = biases,
  effect_direction = c("direct", "inverse")
)
```

<br>

## Suitability | Direct Bias | Strict = FALSE

```{r}
# --- Apply Bias & Sample ---
app_suit_dir <- apply_bias(prepared_bias = prep_biases, 
                           prediction = pred, 
                           prediction_layer = "suitability", 
                           effect_direction = "direct")

occ_suit_dir <- sample_biased_data(n_occ = 100, 
                                   prediction = app_suit_dir, 
                                   prediction_layer = "suitability_biased_direct", 
                                   seed = 123, 
                                   strict = FALSE)

occ_suit_dir_env <- terra::extract(bios,
                                   occ_suit_dir[, c(1:2)]) |>
  cbind(occ_suit_dir[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(1, 2), 
               pch = ".", 
               col_bg = "#9a9797",
               xlab = "Bio1", 
               ylab = "Bio12",
               main = "Suitability | Direct Bias (1v12)")
add_data(occ_suit_dir_env, 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_1",
         y = "bio_12", 
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(2, 3), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15", 
               main = "Suitability | Direct Bias (12v15)")
add_data(occ_suit_dir_env, 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15", 
         pts_col = "red",
         pch = 15)
```

<br>

## Suitability Truncated | Inverse Bias | Strict = TRUE

```{r}
# --- Apply Bias & Sample ---
app_suit_trunc_inv <- apply_bias(prepared_bias = prep_biases, 
                                 prediction = pred, 
                                 prediction_layer = "suitability_trunc", 
                                 effect_direction = "inverse")

occ_suit_trunc_inv <- sample_biased_data(n_occ = 100, 
                                         prediction = app_suit_trunc_inv, 
                                         prediction_layer = "suitability_trunc_biased_inverse", 
                                         strict = TRUE)

occ_suit_trunc_inv_env <- terra::extract(bios, 
                                         occ_suit_trunc_inv[, c(1:2)]) |>
  cbind(occ_suit_trunc_inv[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(1, 2), 
               pch = ".", 
               col_bg = "#9a9797",
               xlab = "Bio1", 
               ylab = "Bio12", 
               main = "Suit Truncated | Inverse Bias (1v12)")
add_data(occ_suit_trunc_inv_env,
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "red", 
         pch = 15)

plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(2, 3),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15", 
               main = "Suit Truncated | Inverse Bias (12v15)")
add_data(occ_suit_trunc_inv_env,
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_12",
         y = "bio_15", 
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis | Direct Bias | Strict = FALSE

```{r}
# --- Apply Bias & Sample ---
app_maha_dir <- apply_bias(prepared_bias = prep_biases, 
                           prediction = pred, 
                           prediction_layer = "Mahalanobis",
                           effect_direction = "direct")

occ_maha_dir <- sample_biased_data(n_occ = 100, 
                                   prediction = app_maha_dir, 
                                   prediction_layer = "Mahalanobis_biased_direct", 
                                   strict = FALSE)

occ_maha_dir_env <- terra::extract(bios, 
                                   occ_maha_dir[, c(1:2)]) |> 
  cbind(occ_maha_dir[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 
plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(1, 2), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1", 
               ylab = "Bio12", 
               main = "Mahalanobis | Direct Bias (1v12)")
add_data(occ_maha_dir_env, 
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_1",
         y = "bio_12", 
         pts_col = "red", 
         pch = 15)

plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(2, 3), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15",
               main = "Mahalanobis | Direct Bias (12v15)")
add_data(occ_maha_dir_env,
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis Truncated | Inverse Bias | Strict = TRUE

```{r}
# --- Apply Bias & Sample ---
app_maha_trunc_inv <- apply_bias(prepared_bias = prep_biases, 
                                 prediction = pred, 
                                 prediction_layer = "Mahalanobis_trunc", 
                                 effect_direction = "inverse")

occ_maha_trunc_inv <- sample_biased_data(n_occ = 100, 
                                         prediction = app_maha_trunc_inv, 
                                         prediction_layer = "Mahalanobis_trunc_biased_inverse", 
                                         strict = TRUE)

occ_maha_trunc_inv_env <- terra::extract(bios, 
                                         occ_maha_trunc_inv[, c(1:2)]) |>
  cbind(occ_maha_trunc_inv[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 
plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]), 
               dim = c(1, 2), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1", 
               ylab = "Bio12", 
               main = "Maha Truncated | Inverse Bias (1v12)")
add_data(occ_maha_trunc_inv_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_1", 
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell, 
               background = as.data.frame(bios[[vars]]), 
               dim = c(2, 3), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15",
               main = "Maha Truncated | Inverse Bias (12v15)")
add_data(occ_maha_trunc_inv_env, 
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)), 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "red",
         pch = 15)

# Clear the plot space when done
dev.off()
```

<br>

# 5. Virtual Species Sampling: `sample_virtual_data()`

**Objective:** To directly generate sampled occurrences from a simulated virtual species object.

**Purpose:** Virtual species are highly useful for controlled experiments in quantitative ecology. This function streamlines the workflow by allowing you to extract occurrence points directly from your virtual species models without manually extracting and converting the underlying rasters first.

This function inherently supports the same weighting rules (`centroid`, `edge`, `random`) and methods (`suitability`, `mahalanobis`) as `sample_data()`.

## Example: Sampling from a Virtual Species Object

```{r}
virtual_range <- data.frame(bio_1 = c(20, 29),
                            bio_12 = c(500, 4000),
                            bio_15 = c(30, 100))

virtual_ell <- build_ellipsoid(range = virtual_range)

# Predict suitability
virtual_pred <- predict(virtual_ell,
                newdata = bios,
                include_mahalanobis = TRUE,
                include_suitability = TRUE,
                suitability_truncated = TRUE,
                mahalanobis_truncated = TRUE)
```

<br>

## Suitability | Centroid | Strict = FALSE

```{r}
# --- Sampling ---
occ_virt_suit_cent <- sample_virtual_data(
  n_occ = 100,
  object = virtual_ell,
  virtual_prediction = virtual_pred,
  prediction_layer = "suitability",
  sampling = "centroid",
  method = "suitability",
  seed = 123,
  strict = FALSE
)

# Extract environmental data
virt_suit_cent_env <- terra::extract(bios, 
                                     occ_virt_suit_cent[, c(1:2)]) |> 
  cbind(occ_virt_suit_cent[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 

# Plot 1: Bio1 vs Bio12
plot_ellipsoid(virtual_ell, 
               dim = c(1, 2), 
               pch = ".",
               col_bg = "#9a9797", 
               xlab = "Bio1",
               ylab = "Bio12", 
               main = "Virtual Suitability | Centroid")
add_data(virt_suit_cent_env, 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)), 
         x = "bio_1", 
         y = "bio_12",
         pts_col = "red",
         pch = 15)

# Plot 2: Bio12 vs Bio15
plot_ellipsoid(virtual_ell,
               dim = c(2, 3), 
               pch = ".",
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15",
               main = "Virtual Suitability | Centroid")
add_data(virt_suit_cent_env, 
         x = "bio_12", 
         y = "bio_15",
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)), 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "red",
         pch = 15)
```

<br>

## Suitability Truncated | Random | Strict = TRUE

```{r}
# --- Sampling ---
occ_virt_suit_trunc_rand <- sample_virtual_data(
  n_occ = 100,
  object = virtual_ell,
  virtual_prediction = virtual_pred,
  prediction_layer = "suitability_trunc",
  sampling = "random",
  method = "suitability",
  seed = 123,
  strict = TRUE
)

# Extract environmental data
virt_suit_trunc_rand_env <- terra::extract(bios, 
                                           occ_virt_suit_trunc_rand[, c(1:2)]) |>
  cbind(occ_virt_suit_trunc_rand[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 

# Plot 1: Bio1 vs Bio12
plot_ellipsoid(virtual_ell, 
               dim = c(1, 2), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Virtual Suit Trunc | Random")
add_data(virt_suit_trunc_rand_env, 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)), 
         x = "bio_1", 
         y = "bio_12",
         pts_col = "red",
         pch = 15)

# Plot 2: Bio12 vs Bio15
plot_ellipsoid(virtual_ell, 
               dim = c(2, 3),
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12",
               ylab = "Bio15", 
               main = "Virtual Suit Trunc | Random")
add_data(virt_suit_trunc_rand_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)), 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis | Edge | Strict = FALSE

```{r}
# --- Sampling ---
occ_virt_maha_edge <- sample_virtual_data(
  n_occ = 100,
  object = virtual_ell,
  virtual_prediction = virtual_pred,
  prediction_layer = "Mahalanobis",
  sampling = "edge",
  method = "mahalanobis",
  seed = 123,
  strict = FALSE
)

# Extract environmental data
virt_maha_edge_env <- terra::extract(bios, 
                                     occ_virt_maha_edge[, c(1:2)]) |>
  cbind(occ_virt_maha_edge[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 

# Plot 1: Bio1 vs Bio12
plot_ellipsoid(virtual_ell, 
               dim = c(1, 2), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1", 
               ylab = "Bio12", 
               main = "Virtual Mahalanobis | Edge")
add_data(virt_maha_edge_env,
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)), 
         x = "bio_1",
         y = "bio_12", 
         pts_col = "red",
         pch = 15)

# Plot 2: Bio12 vs Bio15
plot_ellipsoid(virtual_ell, dim = c(2, 3), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15",
               main = "Virtual Mahalanobis | Edge")
add_data(virt_maha_edge_env, 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)), 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "red", 
         pch = 15)
```

<br>

## Mahalanobis Truncated | Centroid | Strict = TRUE

```{r}
# --- Sampling ---
occ_virt_maha_trunc_cent <- sample_virtual_data(
  n_occ = 100,
  object = virtual_ell,
  virtual_prediction = virtual_pred,
  prediction_layer = "Mahalanobis_trunc",
  sampling = "centroid",
  method = "mahalanobis",
  seed = 123,
  strict = TRUE
)

# Extract environmental data
virt_maha_trunc_cent_env <- terra::extract(bios, 
                                           occ_virt_maha_trunc_cent[, c(1:2)]) |> 
  cbind(occ_virt_maha_trunc_cent[3])

# --- Plotting ---
par(mfrow = c(1, 2)) 

# Plot 1: Bio1 vs Bio12
plot_ellipsoid(virtual_ell, 
               dim = c(1, 2), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio1", 
               ylab = "Bio12", 
               main = "Virt Maha Trunc | Centroid")
add_data(virt_maha_trunc_cent_env, 
         x = "bio_1", 
         y = "bio_12", 
         pts_col = "orange", 
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)), 
         x = "bio_1",
         y = "bio_12",
         pts_col = "red", 
         pch = 15)

# Plot 2: Bio12 vs Bio15
plot_ellipsoid(virtual_ell, 
               dim = c(2, 3), 
               pch = ".", 
               col_bg = "#9a9797", 
               xlab = "Bio12", 
               ylab = "Bio15", 
               main = "Virt Maha Trunc | Centroid")
add_data(virt_maha_trunc_cent_env, 
         x = "bio_12", 
         y = "bio_15", 
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)), 
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)

# Clear the plot space when done
dev.off()
```
